---
title: "Data Manipulation"
author: "Hongtong Lin"
date: "2026-09-24"
output: github_document
---
The file is for doing data manipulation

```{r load library, message = F}
library(tidyverse)
library(dplyr)
```


```{r}
litters_df = 
  read_csv(
    "data/FAS_litters.csv",
    na = c("", "NA", ".")
  )

litters_df = janitor::clean_names(litters_df)

pups_df = 
  read.csv(
    "data/FAS_pups.csv",
    skip = 3,
    na = c("", "NA", ".")
  )

pups_df = janitor::clean_names(pups_df)
```
## Select some variables

```{r}
select(litters_df, group, litter_number)

select(litters_df, group, gd_of_birth)

select(litters_df, -group)

select(litters_df, starts_with("gd"))

select(litters_df, GROUP = group, everything())

select(litters_df, LITTER = litter_number, everything())

select(litters_df, -starts_with("gd"))
```
  
  
a couple of alternatives  

```{r}
rename(litters_df, GROUP = group) #rename col name of group to GROUP

relocate(litters_df, litter_number, gd0_weight) #put columns in the new order, put litter_number and gd0_weight to the very front of the data frame
```

In the pups data, select litter number, sex, and PD ears  

```{r}
select(pups_df, litter_number, sex, pd_ears)
```
  
## filter some rows  

```{r filtering, message=F}
filter(litters_df, gd_of_birth == 20) # the second clause is a logic factor

filter(litters_df, pups_born_alive >5)

filter(litters_df, pups_born_alive != 6)

filter(litters_df, group %in% c("Con7", "Con8"))

filter(litters_df, group %in% c("Con7", "Con8"), pups_born_alive <7)

drop_na(litters_df, gd0_weight)
```

In the pups dataset, keep only rows where PD walk is lees than 11 and sex is 2  

```{r}
filter(pups_df, pd_walk <11, sex ==2)
```
  
  
## `mutate` to change columns  

```{r}
mutate(litters_df, wt_gain = gd18_weight - gd0_weight)
```
  
```{r}
mutate(
  litters_df, 
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)
```
  
In the pups dataset, create a variable that subtracts 7 from PD walk.  
```{r}
mutate(
  pups_df, 
  pd_walk_sbt = pd_walk - 7
)
```
  
  
## `arrange` the data  

```{r}
arrange(litters_df, pups_born_alive)  
arrange(litters_df, desc(pups_born_alive))
```
  
  
## Do multiple steps, use Piping. cmd+shift+M  

```{r}
litters_df = 
  read_csv("data/FAS_litters.csv", na = c("", "NA", ".")) |> #"and then", cmd+shift+M
  janitor::clean_names() |> 
  select(group, starts_with("gd")) |> 
  drop_na() |> 
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
  )

```
  
Load pups, clean names, drop missing, keep litter number and pd variables, and add pd walk-7  
```{r}
pups_df = 
  read_csv("data/FAS_pups.csv", skip =3, na = c("", "NA", ".")) |> 
  janitor::clean_names() |> 
  drop_na() |> 
  select(litter_number, starts_with("pd")) |> 
  mutate(pd_walk_minus_7=pd_walk-7 )
```


